---
title: "ggtree tutorial"
author: Priyamvada Daga (psd2127)
format:
  html:
    embed-resources: true
execute: 
  echo: true
---

## Introduction + Motivation

The `ggtree` package is useful for visualizing and annotating phylogenetic trees. It's built on top of `ggplot2` and has various functions to customize and enhance tree graphics. 

The driving factor is the difficulties I've faced plotting trees, specifically annotated trees, in software like Python. R, via ggtree, provides a neat way of plotting phylogenetic trees, which are a great data visualization tool in the bioinformatics domain.


## Basic features
```{r}
library(ape)       # reading, writing, plotting, and manipulating phylogenetic trees
library(ggplot2)
library(ggtree)
library(dplyr)
```

```{r}
t <- rtree(20)

ggtree(t)
ggtree(t, color="salmon", size=1, linetype="dashed")
ggtree(t, branch.length="none")
```


## Example
```{r}
newick <- "(((((Ruminococcus:1,(prausnitzii:1)Faecalibacterium:1,Oscillospira:1)Ruminococcaceae:1,(SMB53:1,Clostridium:1)Clostridiaceae:1,(Lachnospira:1,(faecis:1)Roseburia:1,Dorea:1,Blautia:1,(gnavus:1)[Ruminococcus]:1,Anaerostipes:1,(eutactus:1)Coprococcus:1)Lachnospiraceae:1,Peptostreptococcaceae:1,(Peptococcus:1)Peptococcaceae:1)Clostridiales:1)Clostridia:1,((((ruminis:1)Lactobacillus:1)Lactobacillaceae:1,(Lactococcus:1,Streptococcus:1)Streptococcaceae:1)Lactobacillales:1)Bacilli:1,(((Bulleidia:1,(dolichum:1)[Eubacterium]:1)Erysipelotrichaceae:1)Erysipelotrichales:1)Erysipelotrichi:1)Firmicutes:1,((((Erwinia:1)Enterobacteriaceae:1)Enterobacteriales:1,(((parainfluenzae:1)Haemophilus:1,Actinobacillus:1)Pasteurellaceae:1)Pasteurellales:1,((Acinetobacter:1)Moraxellaceae:1)Pseudomonadales:1)Gammaproteobacteria:1,(((Sutterella:1)Alcaligenaceae:1)Burkholderiales:1)Betaproteobacteria:1)Proteobacteria:1,(((((ovatus:1,uniformis:1)Bacteroides:1)Bacteroidaceae:1,((copri:1)Prevotella:1)Prevotellaceae:1,((distasonis:1)Parabacteroides:1)Porphyromonadaceae:1,S24-7:1,[Barnesiellaceae]:1)Bacteroidales:1)Bacteroidia:1)Bacteroidetes:1,((RF39:1)Mollicutes:1)Tenericutes:1,(((((lenta:1)Eggerthella:1)Coriobacteriaceae:1)Coriobacteriales:1)Coriobacteriia:1)Actinobacteria:1)Bacteria:1;"
# newick <- gsub("\\[|\\]", "", newick)
tree <- read.tree(text = newick)

color_df <- read.csv("colors.csv")
color_df <- color_df %>%
  mutate(color = ifelse(color == "Mixed", "#808080", color))

tree_data <- as_tibble(tree)

tree_data <- tree_data %>%
  left_join(color_df, by = c("label" = "id"))
```

```{r}
#| fig-width: 14
#| fig-height: 8
ggtree(tree, layout="fan") %<+% tree_data +
  geom_point(aes(color = coalesce(color, "#808080"), size = coalesce(weight, 1)), shape = 16, alpha = 0.5) +
  geom_tree(aes(color = coalesce(color, "#808080")), layout="fan") +
  geom_tiplab(aes(color = coalesce(color, "#808080")), size = 3, hjust=0.05, vjust = -0.3) +
  geom_nodelab(aes(label = label, color = coalesce(color, "#808080")), size = 3, hjust = 0, vjust = -0.3) +
  theme_minimal() +
  scale_color_identity() +
  scale_x_continuous(
    breaks = 0:6,
    labels = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
  ) +
  scale_size_continuous(
    guide = "none"
  )
```
